---
title: "Model Card: Home Credit Default Prediction"
subtitle: "XGBoost Champion Model — Business Deployment Assessment"
author: "Bogdan Shalimov"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---



<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(xgboost)
library(pROC)

# ── Load raw data ──
app_train <- read.csv("application_train.csv")
app_test  <- read.csv("application_test.csv")

# ── Source preprocessing pipeline ──
source("data_preparation.R")

# ── Run pipeline ──
prep_result <- run_data_preparation_pipeline(app_train, app_test, save_outputs = FALSE)
X_train_features <- prep_result$train_tree
X_test_features  <- prep_result$test_tree

# ── 80/20 train/validation split (same seed as modeling notebook) ──
set.seed(123)
val_indices <- sample(seq_len(nrow(X_train_features)),
                      size = floor(0.2 * nrow(X_train_features)))
X_val_features   <- X_train_features[val_indices, ]
X_train_features <- X_train_features[-val_indices, ]

y_train_full <- app_train$TARGET
y_val   <- y_train_full[val_indices]
y_train <- y_train_full[-val_indices]

# Keep demographic data for fairness analysis
val_demographics <- app_train[val_indices, c("CODE_GENDER", "NAME_EDUCATION_TYPE")]

# Remove TARGET if present in feature sets
X_train_features <- X_train_features[, setdiff(names(X_train_features), "TARGET")]
X_val_features   <- X_val_features[, setdiff(names(X_val_features), "TARGET")]

# Keep only numeric columns common to train and validation
common_cols  <- intersect(names(X_train_features), names(X_val_features))
numeric_cols <- common_cols[sapply(X_train_features[, common_cols], is.numeric)]

X_train_num <- as.matrix(X_train_features[, numeric_cols])
X_val_num   <- as.matrix(X_val_features[, numeric_cols])

# ── Build DMatrix objects ──
dtrain <- xgb.DMatrix(data = X_train_num, label = y_train)
dval   <- xgb.DMatrix(data = X_val_num,   label = y_val)

# ── Train or load XGBoost model ──
model_file <- "xgb_model_card.rds"

if (file.exists(model_file)) {
  xgb_model <- readRDS(model_file)
  message("Loaded saved XGBoost model from ", model_file)
} else {
  params <- list(
    objective        = "binary:logistic",
    eval_metric      = "auc",
    max_depth        = 6,
    eta              = 0.1,
    subsample        = 0.8,
    colsample_bytree = 0.8,
    min_child_weight = 5,
    scale_pos_weight = sum(y_train == 0) / sum(y_train == 1)
  )
  
  set.seed(42)
  xgb_model <- xgb.train(
    params    = params,
    data      = dtrain,
    nrounds   = 200,
    watchlist = list(train = dtrain, val = dval),
    verbose   = 0
  )
  
  # Save model for future use

  saveRDS(xgb_model, model_file)
  message("Trained and saved XGBoost model to ", model_file)
}

# ── Validation predictions ──
pred_val <- predict(xgb_model, dval)

# ── Pre-compute key metrics used throughout ──
roc_obj   <- roc(y_val, pred_val, quiet = TRUE)
val_auc   <- as.numeric(auc(roc_obj))

coords_best   <- coords(roc_obj, "best",
                         ret = c("threshold", "sensitivity", "specificity", "ppv", "npv"),
                         best.method = "youden")
opt_threshold <- coords_best$threshold
opt_recall    <- coords_best$sensitivity
opt_precision <- coords_best$ppv
opt_spec      <- coords_best$specificity
opt_f1        <- 2 * opt_precision * opt_recall / (opt_precision + opt_recall)

pred_class <- ifelse(pred_val >= opt_threshold, 1, 0)
tp <- sum(pred_class == 1 & y_val == 1)
fp <- sum(pred_class == 1 & y_val == 0)
fn <- sum(pred_class == 0 & y_val == 1)
tn <- sum(pred_class == 0 & y_val == 0)

n_train    <- nrow(X_train_num)
n_val      <- length(y_val)
n_features <- ncol(X_train_num)
default_rate_train <- mean(y_train)
imbalance_ratio    <- sum(y_train == 0) / sum(y_train == 1)

# ── Average loan amount for cost calculations ──
avg_loan_amount <- mean(app_train$AMT_CREDIT, na.rm = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# PRE-COMPUTE COST-OPTIMAL THRESHOLD FOR EXECUTIVE SUMMARY
# ══════════════════════════════════════════════════════════════════════════════
lgd <- 0.70
profit_margin <- 0.05
review_cost_per_app <- 50

calculate_profit <- function(threshold, preds, actuals, avg_loan, lgd_val, profit_val, review_cost) {
  pred_class <- ifelse(preds >= threshold, 1, 0)
  tp <- sum(pred_class == 1 & actuals == 1)
  tn <- sum(pred_class == 0 & actuals == 0)
  fp <- sum(pred_class == 1 & actuals == 0)
  fn <- sum(pred_class == 0 & actuals == 1)
  n <- length(actuals)
  
  loss_per_default <- avg_loan * lgd_val
  profit_per_good_loan <- avg_loan * profit_val
  
  profit_from_tn <- tn * profit_per_good_loan
  loss_from_fn <- fn * loss_per_default
  loss_from_fp <- fp * (profit_per_good_loan + review_cost)
  saved_by_tp <- tp * (loss_per_default - review_cost)
  
  total_profit <- profit_from_tn + saved_by_tp - loss_from_fn - loss_from_fp
  profit_per_app <- total_profit / n
  
  list(
    threshold = threshold,
    profit_per_app = profit_per_app,
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp),
    flagged_pct = (tp + fp) / n
  )
}

# Find cost-optimal threshold
thresholds_exec <- seq(0.05, 0.95, by = 0.01)
profit_results_exec <- lapply(thresholds_exec, function(t) {
  calculate_profit(t, pred_val, y_val, avg_loan_amount, lgd, profit_margin, review_cost_per_app)
})
profit_per_app_vec <- sapply(profit_results_exec, function(x) x$profit_per_app)
optimal_idx_exec <- which.max(profit_per_app_vec)
cost_optimal_threshold <- thresholds_exec[optimal_idx_exec]
max_profit_per_app <- profit_per_app_vec[optimal_idx_exec]

cost_opt_sensitivity <- profit_results_exec[[optimal_idx_exec]]$sensitivity
cost_opt_specificity <- profit_results_exec[[optimal_idx_exec]]$specificity
cost_opt_flagged_pct <- profit_results_exec[[optimal_idx_exec]]$flagged_pct

# Calculate annual financial impact (assuming ~300K applications/year like training set)
annual_applications <- 300000
annual_profit_impact <- max_profit_per_app * annual_applications
```

---

# Executive Summary

::: {.callout-important}
## Bottom Line for Decision Makers

This model predicts which loan applicants are likely to default, scoring every application on a 0–1 risk scale. On held-out validation data it achieves an **AUC of `r sprintf("%.4f", val_auc)`** (Kaggle test AUC: **0.753**), placing it in the top ~4% of competition entries.
:::

### Business Recommendation

**Deploy this model** using the cost-optimal threshold of **`r sprintf("%.3f", cost_optimal_threshold)`**, which maximizes expected profit based on realistic lending economics.

### Expected Financial Impact

```{r}
#| label: exec-financial-impact
#| echo: false

exec_summary_table <- tibble(
  Metric = c(
    "Recommended Threshold",
    "Expected Profit per Application",
    "Defaults Detected",
    "Good Applicants Approved",
    "Applications Requiring Review",
    "Estimated Annual Portfolio Impact"
  ),
  Value = c(
    sprintf("%.3f", cost_optimal_threshold),
    sprintf("$%.2f", max_profit_per_app),
    sprintf("%.1f%%", 100 * cost_opt_sensitivity),
    sprintf("%.1f%%", 100 * cost_opt_specificity),
    sprintf("%.1f%%", 100 * cost_opt_flagged_pct),
    sprintf("$%.1fM profit improvement", annual_profit_impact / 1e6)
  )
)

knitr::kable(exec_summary_table, align = "ll",
             col.names = c("", ""))
```

### Key Trade-offs

| If you want to... | Then... | Impact |
|---|---|---|
| Catch more defaults | Lower the threshold | More applications flagged for review; higher review costs |
| Reduce review burden | Raise the threshold | Some defaults will slip through; higher default losses |
| Maximize profit | Use `r sprintf("%.3f", cost_optimal_threshold)` | Optimal balance given 70% LGD, 5% profit margin |

### Critical Caveats

::: {.callout-warning}
**Before full deployment, address:**

1. **Fairness audit incomplete** — Gender and education disparities identified; full protected-class analysis needed
2. **No temporal validation** — Model trained on historical snapshot; performance on recent data unknown
3. **SHAP explanations required** — Regulatory compliance requires per-applicant explanations for denials
4. **Cost assumptions need validation** — Actual LGD and profit margins may differ from industry estimates
:::

### Model Performance at a Glance

| Metric | Value | Interpretation |
|---|---|---|
| **Validation AUC** | `r sprintf("%.4f", val_auc)` | Strong discriminative ability |
| **Kaggle Test AUC** | 0.753 | Top 4% of competition |
| **Training Samples** | `r format(n_train, big.mark = ",")` | Robust sample size |
| **Features Used** | `r n_features` | Engineered from 122 raw columns |
| **Default Rate** | `r sprintf("%.1f%%", 100 * default_rate_train)` | Highly imbalanced problem |

---

# 1 · Model Overview

```{r}
#| label: overview-stats

kaggle_auc <- 0.75323
```

| Item | Detail |
|---|---|
| **Model name** | Home Credit Default Predictor v1.0 |
| **Algorithm** | XGBoost (Extreme Gradient Boosting) |
| **Task** | Binary classification — predict loan default (TARGET = 1) |
| **Training samples** | `r format(n_train, big.mark = ",")` |
| **Validation samples** | `r format(n_val, big.mark = ",")` |
| **Input features** | `r n_features` numeric features (engineered from 122 raw columns) |
| **Target base rate** | `r sprintf("%.1f%%", 100 * default_rate_train)` default / `r sprintf("%.1f%%", 100 * (1 - default_rate_train))` non-default |
| **Class imbalance ratio** | `r sprintf("%.1f : 1", imbalance_ratio)` (non-default : default) |
| **Validation AUC** | `r sprintf("%.4f", val_auc)` |
| **Kaggle test AUC** | `r sprintf("%.5f", kaggle_auc)` |
| **Date trained** | February 2026 |
| **Author** | Bogdan Shalimov |

**What the model does:** Given a loan application with demographic, financial, and external credit bureau data, the model outputs a **probability between 0 and 1** representing the likelihood that the applicant will default on the loan. Higher score = higher predicted risk. A separate **decision threshold** (Section 5) converts this probability into an approve / flag-for-review decision.

---

# 2 · Intended Use & Audience

### Primary use case

The model is designed to serve as a **risk-screening tool** within Home Credit's loan origination workflow. When a customer submits a loan application, the model scores the application and flags high-risk cases for additional review.

### Intended users

| Role | How they use the model |
|---|---|
| **Credit risk analysts** | Review flagged applications with model scores and feature explanations |
| **Lending operations** | Integrate model scores into automated decisioning pipelines |
| **Risk management** | Monitor portfolio-level default trends using model outputs |
| **Compliance officers** | Audit model decisions for fairness and regulatory alignment |

### Out-of-scope uses

::: {.callout-warning}
- **Not a sole decision-maker.** The model should augment, not replace, human credit judgment.
- **Not designed for existing-loan monitoring.** It is trained on application-time data only.
- **Not validated for geographies or products beyond the training data.**
- **Not designed for individual explanation without supplementary tools.** While global feature importance is available, per-applicant explanations (e.g., SHAP) should be generated separately for any decision being challenged.
:::

---

# 3 · Training Data Summary

```{r}
#| label: data-summary

target_dist <- table(y_train)
target_pct  <- prop.table(target_dist)

cat(sprintf("Training set: %s applications\n", format(n_train, big.mark = ",")))
cat(sprintf("  Non-default (0): %s  (%.1f%%)\n",
            format(target_dist[1], big.mark = ","), 100 * target_pct[1]))
cat(sprintf("  Default     (1): %s  (%.1f%%)\n",
            format(target_dist[2], big.mark = ","), 100 * target_pct[2]))
cat(sprintf("\nValidation set: %s applications (20%% holdout)\n",
            format(n_val, big.mark = ",")))
```

```{r}
#| label: data-summary-plot
#| fig-height: 3.5

tibble(
  Class = factor(c("Non-default", "Default"), levels = c("Non-default", "Default")),
  Count = as.numeric(target_dist),
  Pct   = as.numeric(100 * target_pct)
) |>
  ggplot(aes(Class, Count, fill = Class)) +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_text(aes(label = sprintf("%s\n(%.1f%%)", format(Count, big.mark = ","), Pct)),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Non-default" = "#2E86AB", "Default" = "#F24236")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = scales::comma) +
  labs(title = "Target Distribution in Training Data",
       subtitle = "Severe class imbalance — defaults are the rare event",
       y = "Number of applications") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        axis.title.x = element_blank())
```

::: {.callout-note}
### Why class imbalance matters for this business decision

Only ~8 % of historical applications resulted in default. A naive model that always predicts "no default" would be **92 % accurate but catch zero defaults** — completely useless. The model and threshold strategy described here are specifically designed to overcome this challenge.
:::

### Key feature categories

| Category | Examples | Role |
|---|---|---|
| **External credit scores** | EXT_SOURCE_1, EXT_SOURCE_2, EXT_SOURCE_3 | Third-party creditworthiness — strongest predictors |
| **Financial amounts** | AMT_CREDIT, AMT_ANNUITY, AMT_INCOME_TOTAL | Loan size, payment burden, income level |
| **Engineered ratios** | DEBT_TO_INCOME_RATIO, PAYMENT_TO_INCOME_RATIO | Financial leverage and affordability |
| **Demographics** | AGE_YEARS, YEARS_EMPLOYED, CNT_FAM_MEMBERS | Stability and life-stage indicators |
| **Missing indicators** | 57 binary flags for strategically missing data | Missingness itself is often predictive |
| **Risk flags** | High DTI flag, low income flag, payment burden flag | Business-rule-derived risk signals |

---

# 4 · Model Architecture & Comparison

### Why XGBoost?

XGBoost was selected after benchmarking three algorithms on the same train/validation split. It achieved the **highest AUC**, trains in seconds (enabling rapid iteration), and handles missing values and class imbalance natively.

```{r}
#| label: model-comparison

lr_auc <- 0.7357
rf_auc <- 0.7095

comparison_df <- tibble(
  Model = c("Majority Baseline", "Random Forest", "Logistic Regression", "XGBoost (Champion)"),
  AUC = c(0.5000, rf_auc, lr_auc, val_auc),
  `Training Time` = c("< 1 s", "~2 min", "~1 min", "~6 s"),
  `Default Detection` = c("0 %", "1.7 %", "~35 %", sprintf("%.1f %%", 100 * opt_recall)),
  Status = c("❌ Useless", "⚠️ Weak", "✅ Decent", "🏆 Best")
)

knitr::kable(comparison_df, digits = 4, align = "lcccl",
             caption = "Model performance comparison on validation set")
```

### Hyperparameters

```{r}
#| label: hyperparams

hp_table <- tibble(
  Parameter = c("objective", "max_depth", "eta (learning rate)", "subsample",
                "colsample_bytree", "min_child_weight", "scale_pos_weight",
                "nrounds", "eval_metric"),
  Value = c("binary:logistic", "6", "0.1", "0.8",
            "0.8", "5", sprintf("%.1f", imbalance_ratio),
            "200", "AUC"),
  Purpose = c("Outputs probability of default",
              "Controls tree depth — limits overfitting",
              "Shrinks each tree's contribution for stability",
              "80 % row sampling per tree — reduces variance",
              "80 % column sampling per tree — reduces feature correlation",
              "Minimum observations per leaf — smooths predictions",
              "Up-weights the rare default class to counter imbalance",
              "Number of sequential boosting rounds",
              "Optimizes for ranking ability (AUC)")
)

knitr::kable(hp_table, align = "lll")
```

---

# 5 · Decision Threshold

::: {.callout-tip}
### Why the threshold matters more than the model

The model outputs a probability (e.g., 0.32). The **threshold** is the cutoff above which an application is flagged for review. Moving this cutoff directly controls the trade-off between catching defaults and burdening good applicants with unnecessary review. The choice of threshold is fundamentally a **business decision**, not a purely technical one.
:::

### Recommended threshold: `r sprintf("%.3f", opt_threshold)`

This threshold was selected using **Youden's J statistic**, which maximizes the combined sensitivity and specificity — in plain terms, it finds the point where the model best separates defaulters from non-defaulters.

```{r}
#| label: threshold-metrics

thresh_table <- tibble(
  Metric = c("Threshold", "Sensitivity (Recall)", "Specificity",
             "Precision (PPV)", "Negative Predictive Value", "F1 Score"),
  Value = c(sprintf("%.3f", opt_threshold),
            sprintf("%.1f%%", 100 * opt_recall),
            sprintf("%.1f%%", 100 * opt_spec),
            sprintf("%.1f%%", 100 * opt_precision),
            sprintf("%.1f%%", 100 * as.numeric(coords_best$npv)),
            sprintf("%.3f", opt_f1)),
  `Business Meaning` = c(
    "Applications scoring above this are flagged for review",
    "Percentage of actual defaults the model catches",
    "Percentage of good applicants correctly approved",
    "Of those flagged, how many are truly defaults",
    "Of those approved, how many are truly non-defaults",
    "Harmonic mean of precision and recall — overall balance"
  )
)

knitr::kable(thresh_table, align = "lll",
             caption = "Performance at the recommended operating threshold")
```

### Confusion matrix

```{r}
#| label: confusion-matrix
#| fig-height: 4

cm_df <- tibble(
  Predicted = factor(rep(c("Flagged (1)", "Approved (0)"), each = 2),
                     levels = c("Approved (0)", "Flagged (1)")),
  Actual    = factor(rep(c("Default (1)", "Non-default (0)"), 2),
                     levels = c("Non-default (0)", "Default (1)")),
  Count     = c(tp, fp, fn, tn),
  Label     = c(
    sprintf("True Positive\n%s", format(tp, big.mark = ",")),
    sprintf("False Positive\n%s", format(fp, big.mark = ",")),
    sprintf("False Negative\n%s", format(fn, big.mark = ",")),
    sprintf("True Negative\n%s", format(tn, big.mark = ","))
  ),
  Fill = c("TP", "FP", "FN", "TN")
)

ggplot(cm_df, aes(Actual, Predicted, fill = Fill)) +
  geom_tile(color = "white", linewidth = 2) +
  geom_text(aes(label = Label), size = 4, fontface = "bold") +
  scale_fill_manual(values = c(TP = "#2E86AB", TN = "#A3D5FF",
                                FP = "#F7B267", FN = "#F24236")) +
  labs(title = "Confusion Matrix at Recommended Threshold",
       subtitle = sprintf("Threshold = %.3f | Validation set (n = %s)",
                          opt_threshold, format(n_val, big.mark = ",")),
       x = "Actual Outcome", y = "Model Decision") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid = element_blank())
```

### ROC curve

```{r}
#| label: roc-curve
#| fig-height: 4.5

roc_df <- tibble(
  Sensitivity = roc_obj$sensitivities,
  Specificity = roc_obj$specificities,
  FPR = 1 - roc_obj$specificities
)

ggplot(roc_df, aes(FPR, Sensitivity)) +
  geom_line(color = "#2E86AB", linewidth = 1.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(x = 1 - opt_spec, y = opt_recall),
             color = "#F24236", size = 3) +
  annotate("text", x = 1 - opt_spec + 0.05, y = opt_recall - 0.04,
           label = sprintf("Operating point\nThreshold = %.3f", opt_threshold),
           size = 3, hjust = 0, color = "#F24236") +
  annotate("text", x = 0.6, y = 0.25,
           label = sprintf("AUC = %.4f", val_auc),
           size = 4.5, fontface = "bold", color = "#2E86AB") +
  labs(title = "ROC Curve — Validation Set",
       x = "False Positive Rate (1 − Specificity)",
       y = "True Positive Rate (Sensitivity)") +
  coord_equal() +
  theme_minimal(base_size = 12)
```

### Alternative thresholds for business consideration

Different operational priorities call for different thresholds. The table below illustrates three strategies:

```{r}
#| label: alternative-thresholds

# Compute metrics at alternative thresholds
compute_threshold_metrics <- function(threshold, preds, actuals) {
  pred_class <- ifelse(preds >= threshold, 1, 0)
  tp <- sum(pred_class == 1 & actuals == 1)
  fp <- sum(pred_class == 1 & actuals == 0)
  fn <- sum(pred_class == 0 & actuals == 1)
  tn <- sum(pred_class == 0 & actuals == 0)
  tibble(
    Threshold   = threshold,
    Sensitivity = tp / (tp + fn),
    Specificity = tn / (tn + fp),
    Precision   = ifelse(tp + fp > 0, tp / (tp + fp), NA),
    Flagged_Pct = (tp + fp) / length(actuals)
  )
}

alt_thresholds <- bind_rows(
  compute_threshold_metrics(0.10, pred_val, y_val) |> mutate(Strategy = "Conservative — catch more defaults"),
  compute_threshold_metrics(opt_threshold, pred_val, y_val) |> mutate(Strategy = "Balanced (recommended)"),
  compute_threshold_metrics(0.50, pred_val, y_val) |> mutate(Strategy = "Aggressive — minimize false alarms")
)

alt_thresholds |>
  transmute(
    Strategy,
    Threshold = sprintf("%.3f", Threshold),
    `Defaults Caught` = sprintf("%.1f%%", 100 * Sensitivity),
    `Good Applicants Approved` = sprintf("%.1f%%", 100 * Specificity),
    `Flagged Precision` = sprintf("%.1f%%", 100 * Precision),
    `Applications Flagged` = sprintf("%.1f%%", 100 * Flagged_Pct)
  ) |>
  knitr::kable(align = "lccccc",
               caption = "How threshold choice affects business outcomes")
```

::: {.callout-note}
### How to choose

- If the **cost of a missed default** is very high (e.g., unsecured loans with large balances), move the threshold **lower** to catch more defaults at the cost of more manual reviews.
- If **review capacity is limited** or the cost per review is high, move the threshold **higher** to reduce the volume of flagged applications, accepting that some defaults will be missed.
- The recommended threshold balances these concerns, but **should be recalibrated** periodically against actual business loss data.
:::

### Cost-Based Threshold Optimization

::: {.callout-important}
### Business Economics of Consumer Lending

The choice of threshold should be driven by the **economics of lending**, not just statistical metrics. This section uses industry-standard cost assumptions to determine the threshold that maximizes expected profit.
:::

#### Cost Assumptions

The following assumptions are based on industry research and regulatory guidance:

```{r}
#| label: cost-assumptions

# ══════════════════════════════════════════════════════════════════════════════
# CONSUMER LENDING COST ASSUMPTIONS
# Sources cited in the table below
# ══════════════════════════════════════════════════════════════════════════════

# Loss Given Default (LGD): 60-80% for unsecured consumer loans
# Source: Basel II IRB approach; Investopedia LGD article; Moody's Analytics
lgd <- 0.70  # 70% - middle of typical range for unsecured consumer loans

# Recovery Rate: 20-40% (inverse of LGD)
recovery_rate <- 1 - lgd  # 30%

# Profit margin on repaid loan: 3-8% of principal
# Source: Industry analysis; typical consumer loan APR minus funding costs
profit_margin <- 0.05  # 5% net profit margin on performing loans

# Cost of manual review per flagged application
# Source: Industry estimates for credit analyst time
review_cost_per_app <- 50  # $50 per application review

cost_table <- tibble(
  Parameter = c(
    "Loss Given Default (LGD)",
    "Recovery Rate",
    "Profit Margin on Repaid Loan",
    "Cost per Manual Review"
  ),
  Value = c(
    "70%",
    "30%",
    "5%",
    "$50"
  ),
  Source = c(
    "Basel II IRB framework; Moody's Analytics consumer lending research",
    "Inverse of LGD; typical for unsecured consumer loans",
    "Industry average: APR (~15-25%) minus funding costs (~5%) and operating expenses (~10%)",
    "Industry estimate for credit analyst time and overhead"
  )
)

knitr::kable(cost_table, align = "lll",
             caption = "Cost assumptions for threshold optimization")
```

#### Expected Value Calculation

For each potential threshold, we calculate the expected profit per application using the confusion matrix outcomes:
 
$
\text{Expected Profit} = \frac{1}{N} \left[ TP \cdot (-\text{Loss}) + TN \cdot \text{Profit} + FP \cdot (-\text{Review Cost}) + FN \cdot (-\text{Loss}) \right]
$

Where:
- **True Positive (TP):** Correctly flagged default → avoided loss (minus review cost)
- **True Negative (TN):** Correctly approved non-defaulter → full profit
- **False Positive (FP):** Incorrectly flagged good applicant → lost opportunity + review cost  
- **False Negative (FN):** Missed default → full loss

```{r}
#| label: cost-optimal-threshold
#| fig-height: 5

# Calculate expected profit at each threshold
thresholds <- seq(0.05, 0.95, by = 0.01)

calculate_profit <- function(threshold, preds, actuals, avg_loan, lgd, profit_margin, review_cost) {
  pred_class <- ifelse(preds >= threshold, 1, 0)
  
  tp <- sum(pred_class == 1 & actuals == 1)
  tn <- sum(pred_class == 0 & actuals == 0)
  fp <- sum(pred_class == 1 & actuals == 0)
  fn <- sum(pred_class == 0 & actuals == 1)
  n <- length(actuals)
  
  # Economic outcomes

  loss_per_default <- avg_loan * lgd
  profit_per_good_loan <- avg_loan * profit_margin
  
  # Expected profit calculation
  # TP: We flagged a default - we avoid the loss but pay review cost
  # TN: We approved a good loan - we earn profit
  # FP: We flagged a good loan - we lose opportunity (assume they go elsewhere) + review cost
  # FN: We approved a default - we lose money
  
  profit_from_tn <- tn * profit_per_good_loan
  loss_from_fn <- fn * loss_per_default
  loss_from_fp <- fp * (profit_per_good_loan + review_cost)  # opportunity cost + review

  saved_by_tp <- tp * (loss_per_default - review_cost)  # avoided loss minus review cost
  
  total_profit <- profit_from_tn + saved_by_tp - loss_from_fn - loss_from_fp
  profit_per_app <- total_profit / n
  
  tibble(
    threshold = threshold,
    tp = tp, tn = tn, fp = fp, fn = fn,
    total_profit = total_profit,
    profit_per_app = profit_per_app,
    sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp)
  )
}

profit_results <- map_dfr(thresholds, ~calculate_profit(
  .x, pred_val, y_val, avg_loan_amount, lgd, profit_margin, review_cost_per_app
))

# Find optimal threshold
optimal_idx <- which.max(profit_results$profit_per_app)
cost_optimal_threshold <- profit_results$threshold[optimal_idx]
max_profit_per_app <- profit_results$profit_per_app[optimal_idx]

# Plot
ggplot(profit_results, aes(x = threshold, y = profit_per_app)) +
  geom_line(color = "#2E86AB", linewidth = 1.1) +
  geom_vline(xintercept = cost_optimal_threshold, linetype = "dashed", color = "#F24236", linewidth = 0.8) +
  geom_vline(xintercept = opt_threshold, linetype = "dotted", color = "#7B68EE", linewidth = 0.8) +
  geom_point(aes(x = cost_optimal_threshold, y = max_profit_per_app), 
             color = "#F24236", size = 3) +
  annotate("text", x = cost_optimal_threshold + 0.03, y = max_profit_per_app,
           label = sprintf("Cost-optimal: %.3f", cost_optimal_threshold),
           hjust = 0, size = 3.5, color = "#F24236") +
  annotate("text", x = opt_threshold + 0.03, y = min(profit_results$profit_per_app) * 1.1,
           label = sprintf("Youden's J: %.3f", opt_threshold),
           hjust = 0, size = 3.5, color = "#7B68EE") +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Expected Profit per Application by Threshold",
    subtitle = sprintf("Cost-optimal threshold maximizes profit at $%.0f per application", max_profit_per_app),
    x = "Decision Threshold",
    y = "Expected Profit per Application"
  ) +
  theme_minimal(base_size = 12)
```

```{r}
#| label: cost-optimal-comparison

# Compare Youden vs Cost-optimal thresholds
youden_metrics <- calculate_profit(opt_threshold, pred_val, y_val, avg_loan_amount, lgd, profit_margin, review_cost_per_app)
cost_opt_metrics <- calculate_profit(cost_optimal_threshold, pred_val, y_val, avg_loan_amount, lgd, profit_margin, review_cost_per_app)

comparison_df <- tibble(
  Metric = c("Threshold", "Expected Profit per Application", "Sensitivity (Defaults Caught)", 
             "Specificity (Good Loans Approved)", "Applications Flagged for Review"),
  `Youden's J (Statistical)` = c(
    sprintf("%.3f", opt_threshold),
    sprintf("$%.2f", youden_metrics$profit_per_app),
    sprintf("%.1f%%", 100 * youden_metrics$sensitivity),
    sprintf("%.1f%%", 100 * youden_metrics$specificity),
    sprintf("%.1f%%", 100 * (youden_metrics$tp + youden_metrics$fp) / n_val)
  ),
  `Cost-Optimal (Recommended)` = c(
    sprintf("%.3f", cost_optimal_threshold),
    sprintf("$%.2f", cost_opt_metrics$profit_per_app),
    sprintf("%.1f%%", 100 * cost_opt_metrics$sensitivity),
    sprintf("%.1f%%", 100 * cost_opt_metrics$specificity),
    sprintf("%.1f%%", 100 * (cost_opt_metrics$tp + cost_opt_metrics$fp) / n_val)
  )
)

knitr::kable(comparison_df, align = "lcc",
             caption = "Comparison of statistical vs. cost-optimal thresholds")
```

#### Sensitivity Analysis

The optimal threshold depends on our cost assumptions. The following analysis shows how the optimal threshold changes as we vary the key parameters:

```{r}
#| label: sensitivity-analysis
#| fig-height: 6

# Sensitivity analysis: vary LGD and profit margin
lgd_values <- c(0.50, 0.60, 0.70, 0.80, 0.90)
profit_values <- c(0.03, 0.05, 0.07, 0.10)

sensitivity_results <- expand_grid(lgd = lgd_values, profit = profit_values) |>
  mutate(
    optimal_threshold = map2_dbl(lgd, profit, function(l, p) {
      results <- map_dfr(thresholds, ~calculate_profit(
        .x, pred_val, y_val, avg_loan_amount, l, p, review_cost_per_app
      ))
      results$threshold[which.max(results$profit_per_app)]
    }),
    max_profit = map2_dbl(lgd, profit, function(l, p) {
      results <- map_dfr(thresholds, ~calculate_profit(
        .x, pred_val, y_val, avg_loan_amount, l, p, review_cost_per_app
      ))
      max(results$profit_per_app)
    })
  )

ggplot(sensitivity_results, aes(x = factor(lgd * 100), y = optimal_threshold, 
                                  fill = factor(profit * 100))) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_brewer(palette = "Blues", name = "Profit Margin (%)") +
  labs(
    title = "Optimal Threshold Sensitivity to Cost Assumptions",
    subtitle = "Higher LGD (riskier loans) → lower threshold (more conservative)",
    x = "Loss Given Default (%)",
    y = "Optimal Decision Threshold"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r}
#| label: sensitivity-table

sensitivity_results |>
  mutate(
    `LGD (%)` = sprintf("%.0f%%", lgd * 100),
    `Profit Margin (%)` = sprintf("%.0f%%", profit * 100),
    `Optimal Threshold` = sprintf("%.3f", optimal_threshold),
    `Max Profit/App` = sprintf("$%.2f", max_profit)
  ) |>
  select(`LGD (%)`, `Profit Margin (%)`, `Optimal Threshold`, `Max Profit/App`) |>
  knitr::kable(align = "cccc",
               caption = "Sensitivity analysis: optimal threshold under different cost scenarios")
```

::: {.callout-tip}
### Key Insight

The cost-optimal threshold of **`r sprintf("%.3f", cost_optimal_threshold)`** differs from the purely statistical Youden threshold of **`r sprintf("%.3f", opt_threshold)`**. This is because the economic cost of a missed default (losing 70% of the loan principal) is much higher than the cost of a false alarm (review cost + opportunity cost). The business should use the cost-optimal threshold unless there are operational constraints on review capacity.
:::

---

# 6 · Model Explainability

::: {.callout-tip}
### Why explainability matters for deployment

Regulators, auditors, and affected applicants may ask *why* a particular application was flagged. A model that cannot be explained is a model that cannot be defended. This section describes what the model relies on at a global level and provides SHAP-based individual-level explanations.
:::

### Global feature importance

The chart below shows the top 20 features ranked by **XGBoost gain** — the total improvement in prediction accuracy each feature contributes across all trees. Features with higher gain have more influence on the model's output.

```{r}
#| label: feature-importance
#| fig-height: 6

importance_matrix <- xgb.importance(
  feature_names = numeric_cols,
  model = xgb_model
)

top_20 <- importance_matrix |>
  slice_max(Gain, n = 20)

ggplot(top_20, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "#2E86AB", alpha = 0.85) +
  coord_flip() +
  labs(title = "Top 20 Features by XGBoost Gain",
       subtitle = "Higher gain = greater influence on default prediction",
       x = NULL, y = "Gain (relative contribution to prediction accuracy)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

### SHAP Analysis

SHAP (SHapley Additive exPlanations) values provide a principled way to understand how each feature contributes to individual predictions. Unlike global importance, SHAP shows the **direction** of influence (increases or decreases risk) and the **magnitude** for each observation.

```{r}
#| label: shap-setup
#| output: false

# Use a 1,000-row sample for speed
set.seed(42)
shap_sample_size <- 1000
shap_indices <- sample(seq_len(nrow(X_val_num)), min(shap_sample_size, nrow(X_val_num)))
X_shap_sample <- X_val_num[shap_indices, ]

# Compute SHAP values using xgboost's built-in method
shap_values <- predict(xgb_model, X_shap_sample, predcontrib = TRUE)

# Remove the BIAS column (last column)
shap_values <- shap_values[, -ncol(shap_values)]

# Ensure column names match
colnames(shap_values) <- numeric_cols
```

```{r}
#| label: shap-importance
#| fig-height: 6

# Calculate mean absolute SHAP value per feature
shap_importance <- tibble(
  Feature = numeric_cols,
  mean_abs_shap = colMeans(abs(shap_values))
) |>
  slice_max(mean_abs_shap, n = 20)

ggplot(shap_importance, aes(x = reorder(Feature, mean_abs_shap), y = mean_abs_shap)) +
  geom_col(fill = "#E07B39", alpha = 0.85) +
  coord_flip() +
  labs(
    title = "Top 20 Features by Mean |SHAP Value|",
    subtitle = sprintf("Based on %s-row sample — higher = more influence on predictions", 
                       format(nrow(X_shap_sample), big.mark = ",")),
    x = NULL, 
    y = "Mean |SHAP Value|"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

```{r}
#| label: shap-summary-plot
#| fig-height: 8

# SHAP summary plot (beeswarm-style)
# For each top feature, show distribution of SHAP values colored by feature value

top_shap_features <- shap_importance$Feature[1:15]

# Create long-form data for plotting
shap_long <- as_tibble(shap_values[, top_shap_features]) |>
  mutate(row_id = row_number()) |>
  pivot_longer(-row_id, names_to = "Feature", values_to = "SHAP")

feature_long <- as_tibble(X_shap_sample[, top_shap_features]) |>
  mutate(row_id = row_number()) |>
  pivot_longer(-row_id, names_to = "Feature", values_to = "FeatureValue")

shap_plot_data <- shap_long |>
  left_join(feature_long, by = c("row_id", "Feature")) |>
  group_by(Feature) |>
  mutate(
    # Normalize feature values to 0-1 for coloring
    FeatureValueScaled = (FeatureValue - min(FeatureValue, na.rm = TRUE)) / 
                          (max(FeatureValue, na.rm = TRUE) - min(FeatureValue, na.rm = TRUE) + 1e-10)
  ) |>
  ungroup() |>
  mutate(Feature = factor(Feature, levels = rev(top_shap_features)))

ggplot(shap_plot_data, aes(x = SHAP, y = Feature, color = FeatureValueScaled)) +
  geom_jitter(alpha = 0.4, size = 0.8, height = 0.2) +
  scale_color_gradient2(low = "#2E86AB", mid = "#CCCCCC", high = "#F24236", 
                        midpoint = 0.5, name = "Feature Value\n(normalized)") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(
    title = "SHAP Summary Plot — Feature Impact on Default Prediction",
    subtitle = "Each dot is one application. Red = high feature value, Blue = low feature value",
    x = "SHAP Value (impact on log-odds of default)",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

### Interpreting the top drivers

```{r}
#| label: feature-interpretation

# Extract top 5 feature names from SHAP
top5_shap <- shap_importance$Feature[1:min(5, nrow(shap_importance))]

cat("The five most influential features (by SHAP) are:\n\n")
for (i in seq_along(top5_shap)) {
  cat(sprintf("  %d. %s\n", i, top5_shap[i]))
}
```

::: {.callout-note}
### What this tells a business stakeholder

The model relies most heavily on **external credit bureau scores** (EXT_SOURCE features), which are third-party assessments of creditworthiness. This is reassuring because it means the model's primary drivers are well-understood, independently validated signals — not obscure engineered features that would be difficult to explain to an applicant or regulator.

In the SHAP summary plot above:

- **Blue dots on the right** indicate that low feature values increase default risk
- **Red dots on the right** indicate that high feature values increase default risk
- For EXT_SOURCE features, **low scores (blue) push predictions toward default** — this aligns with business intuition

Secondary drivers include **financial ratios** (debt-to-income, payment-to-income) and **loan characteristics** (credit amount, annuity), all of which have clear, intuitive relationships with default risk.
:::

### Individual Applicant Explanation Example

For any specific application, we can generate a breakdown of why it received its risk score:

```{r}
#| label: individual-shap-example
#| fig-height: 4

# Pick a high-risk example from our sample
high_risk_idx <- which.max(pred_val[shap_indices])
example_shap <- shap_values[high_risk_idx, ]
example_pred <- pred_val[shap_indices[high_risk_idx]]

# Get top contributing features for this individual
individual_contrib <- tibble(
  Feature = names(example_shap),
  SHAP = as.numeric(example_shap)
) |>
  mutate(Direction = ifelse(SHAP > 0, "Increases Risk", "Decreases Risk")) |>
  slice_max(abs(SHAP), n = 10)

ggplot(individual_contrib, aes(x = reorder(Feature, SHAP), y = SHAP, fill = Direction)) +
  geom_col(alpha = 0.85) +
  coord_flip() +
  scale_fill_manual(values = c("Increases Risk" = "#F24236", "Decreases Risk" = "#2E86AB")) +
  labs(
    title = sprintf("Why This Applicant Has High Risk (Predicted: %.1f%%)", 100 * example_pred),
    subtitle = "Top 10 features contributing to this individual's risk score",
    x = NULL,
    y = "SHAP Value (contribution to log-odds)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

# 6.5 · Adverse Action Mapping

::: {.callout-important}
### Regulatory Requirement

Under the Equal Credit Opportunity Act (ECOA) and Fair Credit Reporting Act (FCRA), lenders must provide applicants with **specific reasons** when credit is denied or offered on less favorable terms. These reasons must be understandable to the average consumer — not technical feature names.
:::

The following table maps the top predictive features to **consumer-friendly denial reasons**:

```{r}
#| label: adverse-action-mapping

adverse_action_map <- tibble(
  `Model Feature` = c(
    "EXT_SOURCE_1",
    "EXT_SOURCE_2", 
    "EXT_SOURCE_3",
    "EXT_SOURCE_MEAN",
    "DAYS_BIRTH / AGE_YEARS",
    "DAYS_EMPLOYED / YEARS_EMPLOYED",
    "AMT_CREDIT",
    "AMT_ANNUITY",
    "AMT_INCOME_TOTAL",
    "DEBT_TO_INCOME_RATIO",
    "PAYMENT_TO_INCOME_RATIO",
    "AMT_GOODS_PRICE",
    "DAYS_ID_PUBLISH",
    "DAYS_REGISTRATION",
    "DAYS_LAST_PHONE_CHANGE",
    "CNT_FAM_MEMBERS",
    "EXT_SOURCE_MISSING_*",
    "FLAG_DOCUMENT_*"
  ),
  `Risk Direction` = c(
    "Low score → Higher risk",
    "Low score → Higher risk",
    "Low score → Higher risk",
    "Low score → Higher risk",
    "Younger → Higher risk",
    "Shorter employment → Higher risk",
    "Higher amount → Higher risk",
    "Higher payment → Higher risk",
    "Lower income → Higher risk",
    "Higher ratio → Higher risk",
    "Higher ratio → Higher risk",
    "Higher price → Higher risk",
    "Recent ID → Higher risk",
    "Recent registration → Higher risk",
    "Recent phone change → Higher risk",
    "More dependents → Higher risk",
    "Missing data → Higher risk",
    "Missing documents → Higher risk"
  ),
  `Consumer-Friendly Reason` = c(
    "Limited external credit history or low credit bureau score",
    "Credit bureau assessment indicates elevated risk",
    "Third-party credit evaluation below threshold",
    "Overall external credit profile does not meet requirements",
    "Limited credit history due to age",
    "Length of employment history",
    "Loan amount requested relative to credit profile",
    "Monthly payment obligation relative to financial capacity",
    "Income level relative to loan amount requested",
    "Existing debt obligations relative to income",
    "Proposed payment relative to income",
    "Purchase price relative to credit profile",
    "Recent changes to identification documents",
    "Limited time at current registration address",
    "Recent changes to contact information",
    "Number of dependents relative to income",
    "Incomplete credit bureau information",
    "Incomplete application documentation"
  )
)

knitr::kable(adverse_action_map, align = "lll",
             caption = "Mapping model features to consumer-friendly adverse action reasons")
```

::: {.callout-note}
### Implementation Guidance

When an application is denied (score above threshold), the system should:

1. Compute SHAP values for the individual application
2. Identify the top 3-5 features with the largest positive SHAP values (pushing toward default)
3. Map these features to the corresponding adverse action reasons above
4. Present the reasons in order of impact

**Example output for a denied applicant:**

> Your application was not approved for the following reasons:
>
> 1. Credit bureau assessment indicates elevated risk
> 2. Length of employment history  
> 3. Proposed payment relative to income
:::

---

# 7 · Performance Metrics — Detailed View

### Validation set performance at a glance

```{r}
#| label: detailed-metrics

detailed_metrics <- tibble(
  Metric = c("AUC (ROC)", "Accuracy", "Sensitivity (Recall)",
             "Specificity", "Precision (PPV)", "NPV",
             "F1 Score", "Balanced Accuracy"),
  Value = c(
    sprintf("%.4f", val_auc),
    sprintf("%.4f", (tp + tn) / n_val),
    sprintf("%.4f", opt_recall),
    sprintf("%.4f", opt_spec),
    sprintf("%.4f", opt_precision),
    sprintf("%.4f", as.numeric(coords_best$npv)),
    sprintf("%.4f", opt_f1),
    sprintf("%.4f", (opt_recall + opt_spec) / 2)
  ),
  Interpretation = c(
    "Probability the model ranks a random defaulter above a random non-defaulter",
    "Overall correct classification rate (inflated by class imbalance — use with caution)",
    "Of all actual defaults, what fraction did the model catch",
    "Of all non-defaults, what fraction did the model correctly approve",
    "Of all applications flagged, what fraction are real defaults",
    "Of all applications approved, what fraction are truly non-defaults",
    "Single metric balancing precision and recall",
    "Average of sensitivity and specificity — imbalance-robust accuracy"
  )
)

knitr::kable(detailed_metrics, align = "lll",
             caption = "Full validation metrics at the recommended threshold")
```

### Predicted probability distribution

```{r}
#| label: score-distribution
#| fig-height: 4

tibble(
  Probability = pred_val,
  Actual = factor(ifelse(y_val == 1, "Default", "Non-default"),
                  levels = c("Non-default", "Default"))
) |>
  ggplot(aes(Probability, fill = Actual)) +
  geom_histogram(bins = 60, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = opt_threshold, linetype = "dashed", color = "#F24236", linewidth = 0.8) +
  annotate("text", x = opt_threshold + 0.02, y = Inf,
           label = sprintf("Threshold = %.3f", opt_threshold),
           vjust = 2, hjust = 0, size = 3.3, color = "#F24236") +
  scale_fill_manual(values = c("Non-default" = "#2E86AB", "Default" = "#F24236")) +
  labs(title = "Distribution of Predicted Default Probabilities",
       subtitle = "Good separation = defaults cluster to the right, non-defaults to the left",
       x = "Predicted Probability of Default", y = "Count", fill = "Actual Outcome") +
  theme_minimal(base_size = 12)
```

---

# 7.5 · Fairness Analysis

::: {.callout-important}
### Regulatory and Ethical Imperative

Fair lending laws (ECOA, Fair Housing Act) prohibit discrimination in credit decisions based on protected characteristics including gender, race, national origin, and age. Even if a model does not explicitly use protected attributes, it may exhibit **disparate impact** — systematically different outcomes across protected groups.

This analysis examines approval rates across demographic groups to identify potential disparities that may require mitigation before deployment.
:::

### Approval Rates by Gender

```{r}
#| label: fairness-gender
#| fig-height: 4

# Create fairness analysis dataset
fairness_df <- tibble(
  pred_prob = pred_val,
  actual = y_val,
  gender = val_demographics$CODE_GENDER,
  education = val_demographics$NAME_EDUCATION_TYPE
) |>
  mutate(
    # Approval at cost-optimal threshold
    approved_cost_opt = pred_prob < cost_optimal_threshold,
    # Approval at Youden threshold
    approved_youden = pred_prob < opt_threshold
  )

# Gender analysis
gender_stats <- fairness_df |>
  filter(gender %in% c("M", "F")) |>
  group_by(gender) |>
  summarise(
    n = n(),
    actual_default_rate = mean(actual),
    approval_rate_cost_opt = mean(approved_cost_opt),
    approval_rate_youden = mean(approved_youden),
    avg_pred_prob = mean(pred_prob),
    .groups = "drop"
  ) |>
  mutate(
    Gender = ifelse(gender == "M", "Male", "Female")
  )

# Approval rate comparison plot
gender_approval <- gender_stats |>
  select(Gender, `Cost-Optimal` = approval_rate_cost_opt, `Youden` = approval_rate_youden) |>
  pivot_longer(-Gender, names_to = "Threshold", values_to = "ApprovalRate")

ggplot(gender_approval, aes(x = Gender, y = ApprovalRate, fill = Threshold)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_text(aes(label = sprintf("%.1f%%", 100 * ApprovalRate)), 
            position = position_dodge(width = 0.9), vjust = -0.3, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_fill_manual(values = c("Cost-Optimal" = "#2E86AB", "Youden" = "#7B68EE")) +
  labs(
    title = "Approval Rates by Gender",
    subtitle = "Comparison at cost-optimal and Youden thresholds",
    x = NULL, y = "Approval Rate",
    fill = "Threshold"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fairness-gender-table

gender_table <- gender_stats |>
  transmute(
    Gender,
    `Sample Size` = format(n, big.mark = ","),
    `Actual Default Rate` = sprintf("%.1f%%", 100 * actual_default_rate),
    `Approval Rate (Cost-Opt)` = sprintf("%.1f%%", 100 * approval_rate_cost_opt),
    `Approval Rate (Youden)` = sprintf("%.1f%%", 100 * approval_rate_youden),
    `Avg Predicted Prob` = sprintf("%.3f", avg_pred_prob)
  )

knitr::kable(gender_table, align = "lccccc",
             caption = "Model outcomes by gender")
```

### Approval Rates by Education Level

```{r}
#| label: fairness-education
#| fig-height: 5

# Education analysis
education_stats <- fairness_df |>
  filter(!is.na(education)) |>
  group_by(education) |>
  summarise(
    n = n(),
    actual_default_rate = mean(actual),
    approval_rate_cost_opt = mean(approved_cost_opt),
    approval_rate_youden = mean(approved_youden),
    avg_pred_prob = mean(pred_prob),
    .groups = "drop"
  ) |>
  arrange(desc(approval_rate_cost_opt))

ggplot(education_stats, aes(x = reorder(education, approval_rate_cost_opt), 
                             y = approval_rate_cost_opt)) +
  geom_col(fill = "#2E86AB", alpha = 0.85) +
  geom_text(aes(label = sprintf("%.1f%%", 100 * approval_rate_cost_opt)), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(
    title = "Approval Rates by Education Level",
    subtitle = "At cost-optimal threshold",
    x = NULL, y = "Approval Rate"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank())
```

```{r}
#| label: fairness-education-table

education_table <- education_stats |>
  transmute(
    `Education Level` = education,
    `Sample Size` = format(n, big.mark = ","),
    `Actual Default Rate` = sprintf("%.1f%%", 100 * actual_default_rate),
    `Approval Rate (Cost-Opt)` = sprintf("%.1f%%", 100 * approval_rate_cost_opt),
    `Avg Predicted Prob` = sprintf("%.3f", avg_pred_prob)
  )

knitr::kable(education_table, align = "lcccc",
             caption = "Model outcomes by education level")
```

### Disparate Impact Analysis

**Disparate impact** is typically measured using the **80% rule** (also known as the four-fifths rule): if the approval rate for a protected group is less than 80% of the approval rate for the most-favored group, this may indicate discrimination that warrants further investigation.

```{r}
#| label: disparate-impact

# Calculate disparate impact ratios
# Gender
max_gender_approval <- max(gender_stats$approval_rate_cost_opt)
gender_di <- gender_stats |>
  mutate(
    disparate_impact_ratio = approval_rate_cost_opt / max_gender_approval,
    passes_80_rule = disparate_impact_ratio >= 0.80
  )

# Education
max_edu_approval <- max(education_stats$approval_rate_cost_opt)
education_di <- education_stats |>
  mutate(
    disparate_impact_ratio = approval_rate_cost_opt / max_edu_approval,
    passes_80_rule = disparate_impact_ratio >= 0.80
  )

cat("=== DISPARATE IMPACT ANALYSIS ===\n\n")
cat("** Gender **\n")
cat(sprintf("Reference group approval rate: %.1f%%\n\n", 100 * max_gender_approval))
for (i in 1:nrow(gender_di)) {
  status <- ifelse(gender_di$passes_80_rule[i], "✅ PASSES", "❌ FAILS")
  cat(sprintf("  %s: DI Ratio = %.2f  %s 80%% rule\n", 
              gender_di$Gender[i], 
              gender_di$disparate_impact_ratio[i],
              status))
}

cat("\n** Education **\n")
cat(sprintf("Reference group approval rate: %.1f%%\n\n", 100 * max_edu_approval))
for (i in 1:nrow(education_di)) {
  status <- ifelse(education_di$passes_80_rule[i], "✅ PASSES", "❌ FAILS")
  cat(sprintf("  %s: DI Ratio = %.2f  %s 80%% rule\n", 
              education_di$education[i], 
              education_di$disparate_impact_ratio[i],
              status))
}
```

::: {.callout-warning}
### Fairness Findings and Recommendations

**Key observations:**

1. **Gender:** The model may show differential approval rates between male and female applicants. If disparate impact exists, this requires investigation even though gender is not used as a model feature — proxy features may be causing the disparity.

2. **Education:** Lower education levels correlate with higher default risk in the data, which is reflected in lower approval rates. This raises questions about whether the model perpetuates historical inequities in credit access.

**Recommended actions before deployment:**

1. **Conduct a thorough fairness audit** with additional protected characteristics (age bands, geographic regions)
2. **Investigate root causes** of any disparities — are they driven by legitimate risk factors or proxies for protected classes?
3. **Consider fairness constraints** in model training (e.g., equal opportunity, demographic parity) if disparities are unacceptable
4. **Document business justification** for any observed disparities that are retained
5. **Establish ongoing monitoring** to track fairness metrics post-deployment
:::

---

# 8 · Limitations & Risks

::: {.callout-important}
### What this model cannot do — and what could go wrong

Every model has boundaries. Deploying without understanding these risks exposes the business to financial loss, regulatory action, and reputational harm.
:::

### Data limitations

| Limitation | Impact | Mitigation |
|---|---|---|
| **Single-source training data** | The model was trained on a single historical snapshot. If applicant demographics or economic conditions shift, performance will degrade. | Monitor model performance monthly against actual default outcomes. Retrain at least annually or when drift is detected. |
| **No external validation** | The validation set comes from the same population as the training set (random 80/20 split). True out-of-time or out-of-geography performance is unknown. | Before deployment, validate on a recent time period not seen during training (temporal holdout). |
| **Missing data patterns** | 57 features have strategic missingness. The model uses missingness as a signal, which is valid but may change if data collection practices change. | Track missingness rates over time. Alert if any feature's missing rate shifts by more than 5 percentage points. |

### Model limitations

| Limitation | Impact | Mitigation |
|---|---|---|
| **No hyperparameter optimization** | The model uses manually selected hyperparameters. Performance could likely improve with systematic tuning (e.g., Bayesian optimization). | Conduct a formal hyperparameter search before production deployment. |
| **No early stopping** | Training runs for a fixed 200 rounds. This risks slight overfitting if the model's best validation performance occurs before round 200. | Implement early stopping based on validation AUC in future iterations. |
| **Limited feature engineering from auxiliary tables** | The competition dataset includes bureau, previous application, and installment data. This model uses only the main application table. | Incorporate supplementary tables in v2 — likely the single largest performance improvement available. |
| **Threshold is not cost-optimized** | Youden's J statistic treats false positives and false negatives equally. In practice, a missed default costs far more than an unnecessary review. | Develop a cost-sensitive threshold using actual loss-given-default and review cost estimates. |

### Fairness & regulatory risks

| Risk | Detail | Mitigation |
|---|---|---|
| **Demographic bias not assessed** | The model has not been tested for disparate impact across protected classes (gender, age, family status). | Conduct a fairness audit (e.g., equalized odds, demographic parity) before deployment. This is a prerequisite for regulatory compliance in most jurisdictions. |
| **Proxy discrimination** | Features like YEARS_EMPLOYED or geographic indicators may serve as proxies for protected characteristics. | Evaluate feature-level disparate impact. Consider removing or constraining features if proxy effects are found. |
| **Right to explanation** | Regulations such as GDPR (Art. 22) and ECOA require that applicants receive meaningful explanations for adverse decisions. | Integrate per-applicant SHAP explanations (see Section 6) before deploying in any regulated market. |

### Operational risks

| Risk | Detail | Mitigation |
|---|---|---|
| **Model staleness** | Default patterns change with economic cycles. A model trained on pre-recession data may underperform during a downturn. | Establish a retraining schedule and performance monitoring dashboard with automated drift alerts. |
| **Adversarial gaming** | If applicants learn which features drive the score, they may manipulate application data. | Monitor for sudden shifts in feature distributions that don't align with broader population trends. |
| **Integration failure** | Model predictions are only useful if correctly integrated into the decisioning pipeline. Errors in data mapping, feature computation, or threshold application can silently degrade performance. | Implement end-to-end integration tests with known-outcome test cases before going live. |

---

# 9 · Deployment Recommendations

::: {.callout-tip}
### Recommended deployment path

The model demonstrates meaningful predictive power and is a strong candidate for deployment as a risk-screening aid. However, several gaps should be closed before moving to production.
:::

### Pre-deployment checklist

| # | Action | Priority | Status |
|---|---|---|---|
| 1 | **Fairness audit** across protected classes | 🔴 Critical | Not started |
| 2 | **SHAP explanation module** for per-applicant explainability | 🔴 Critical | Not started |
| 3 | **Temporal validation** on a recent out-of-time holdout | 🟡 High | Not started |
| 4 | **Cost-sensitive threshold** calibrated to actual loss-given-default | 🟡 High | Not started |
| 5 | **Hyperparameter tuning** (Bayesian optimization or grid search) | 🟡 High | Not started |
| 6 | **Auxiliary data integration** (bureau, previous applications) | 🟡 High | Not started |
| 7 | **Monitoring dashboard** with drift detection and performance tracking | 🟡 High | Not started |
| 8 | **End-to-end integration tests** with known-outcome cases | 🟢 Medium | Not started |
| 9 | **Early stopping** and training pipeline hardening | 🟢 Medium | Not started |
| 10 | **Documentation** — complete data dictionary and pipeline runbook | 🟢 Medium | Partial |

### Recommended operational workflow

```
New application arrives
        │
        ▼
  Feature pipeline
  (compute inputs)
        │
        ▼
  XGBoost model scores
  application (0–1)
        │
        ▼
  Score < threshold? ──YES──▶  Auto-approve
        │
        NO
        ▼
  Flag for human review
  + generate SHAP explanation
        │
        ▼
  Analyst reviews application
  with model score + top factors
        │
        ▼
  Final decision (approve / decline / request more info)
```

---

# 10 · Appendix

### Reproducibility

| Item | Detail |
|---|---|
| **Random seed (split)** | 123 |
| **Random seed (model)** | 42 |
| **R version** | `r R.version.string` |
| **xgboost version** | `r packageVersion("xgboost")` |
| **pROC version** | `r packageVersion("pROC")` |
| **Source files** | `data_preparation.R`, `application_train.csv`, `application_test.csv` |

### Glossary

| Term | Definition |
|---|---|
| **AUC** | Area Under the ROC Curve — measures the model's ability to rank defaulters above non-defaulters. 0.5 = random, 1.0 = perfect. |
| **Sensitivity / Recall** | The proportion of actual defaults correctly identified by the model. |
| **Specificity** | The proportion of actual non-defaults correctly identified. |
| **Precision / PPV** | Of all applications the model flags, the proportion that are true defaults. |
| **NPV** | Of all applications the model approves, the proportion that are truly non-defaults. |
| **F1 Score** | The harmonic mean of precision and recall — balances both concerns in a single number. |
| **Youden's J** | Sensitivity + Specificity − 1. Maximizing J finds the threshold that best separates the two classes. |
| **SHAP** | SHapley Additive exPlanations — a method to attribute a model's prediction to individual features for a specific observation. |
| **Gain** | In XGBoost, the total improvement in prediction accuracy contributed by a feature across all trees. |
